---
---

## Split-apply-combine

A very common data manipulation procedure is doing some "group-wise" operations on a dataset and combing the results for each group into a single table. For example, say you need to count the number of individuals *of each species* observed in the winter of 1990.

===

## Grouping

The dplyr function `group_by` begins the process by indicating how the data frame should be split into subsets.

```{r summarize, title="{{ site.handouts[0] }}"}
counts_1990_winter <- animals_1990_winter %>%
    group_by(species_id)
```

===

At this point, nothing has really changed:

```{r}
str(counts_1990_winter)
```

The `group_by` statement does not change any values in the data frame; it only adds attributes to the the original data frame. You can add multiple variables (separated by commas) in `group_by`; each distinct combination of values across these columns defines a different group.
{:.notes}

===

## Summarize

The operation to perform on each species is counting: we need to count how many records are in each group.

```{r summarize_n, title="{{ site.handouts[0] }}"}
counts_1990_winter <- animals_1990_winter %>%
    group_by(species_id) %>%
    summarize(count = n())
```

===

```{r view_summarize}
str(counts_1990_winter)
```

===

The "combine" part of "split-apply-combine" occurs automatically, when the attributes introduced by `group_by` are dropped. You can see attributes either by running the `str()` function on the data frame or by inspecting it in the RStudio *Environment* pane.

===

The function `n()` takes no arguments and returns the number of records in a group. Any function that collapses a vector input to a single output is a suitable function to use within `summarize`.

```{r title="{{ site.handouts[0] }}"}
weight_1990_winter <- animals_1990_winter %>%
    group_by(species_id) %>%
    summarize(avg_weight = mean(weight, na.rm = TRUE))
```

===

```{r}
head(weight_1990_winter)
```

===

## Exercise 3

Write code that returns the average hindfoot length of *Dipodomys merriami* (DM) individuals observed in each month (irrespective of the year). Make sure to exclude *NA* values.

[View solution](#solution-3)
{:.notes}

===

## Transformation of variables

The `mutate` function creates new columns from existing ones. The data frame returned has an additional column for each argument to `mutate`, unless a name is reused. Overwriting an existing column does not generate a warning.

===

The `count` variable just defined, for example, can be used to calculate the proportion of individuals represented by each species.

```{r mutate, title="{{ site.handouts[0] }}"}
prop_1990_winter <- counts_1990_winter %>%
    mutate(prop = count / sum(count))
```

===

```{r}
head(prop_1990_winter)
```
For a concise way to apply the same transformation to multiple columns, check the `mutate_each` function. There is also a `summarize_each` function to perform the same aggregation operation on multiple columns.
{:.notes}

===

Both `mutate` and `summarize` can be used in the "apply" part of a "split-apply-combine" procedure. The difference is that the results are combine into data frames with differing numbers of rows.

===

Question
: How many rows do you expect in the result of a `mutate` operation?

Answer
: {:.fragment} The same number you started with.

===

## Exercise 4

A "pivot table" is a transformation of tidy data into a wide summary table. First, data are summarized by *two* grouping factors, then one of these is "pivoted" into columns. Starting from the `animals` data frame, chain a `group_by` and `summarize` transformation into a [tidyr](){:.rlib} `spread` function to get the number of individuals counted in each month (as three columns) by species (as rows).
